\documentclass[aps,prd,10pt,onecolumn,nofootinbib,amsmath,amssymb]{revtex4-1}

\usepackage[colorlinks=true,breaklinks=true]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{slashed}
\usepackage{xcolor}

\hypersetup{
  urlcolor=blue,
  citecolor=blue,
  linkcolor=blue
}

\newcommand{\dd}{\mathrm{d}}
\newcommand{\vev}[1]{\left\langle #1\right\rangle}

\begin{document}

\title{Excitation and de-excitation rates from a relativistic reactor
antineutrino source}
\author{Working derivation based on Appendix B of \texttt{main\_v4.tex}}
\noaffiliation
\date{\today}
\maketitle

\section{Scope and assumptions}
\label{sec:scope}

This note repeats the excitation and de-excitation calculation of
Appendix~B for a reactor source.  We make the following assumptions.
\begin{enumerate}
  \item The source contains electron antineutrinos with
  \(E\sim\mathrm{MeV}\gg m_i\), so all neutrino masses are neglected in
  external-state kinematics and spinor traces.
  \item The produced antineutrinos have the active, positive-helicity state.
  Contributions from the wrong helicity are of relative order
  \(m_i^2/E^2\).
  \item Only neutral-current scattering from a nuclear spin is included.
  The neutral-current axial coupling \(g_A\) is flavour universal, diagonal
  in the neutrino mass basis, and the nucleus is treated as pointlike on the
  scale of the momentum transfer relevant here.
  \item The nuclear spin is fixed in the sample.  Target recoil and energy
  transfer to phonons are neglected, as requested.  Consequently the
  nuclear spin can absorb momentum without a recoil-energy term in the
  energy-conservation delta function.
  \item The reactor occupation is extremely dilute, so final-state Pauli
  blocking is set to one.
  \item The incident field is first treated locally as a beam of plane
  waves.  A finite core is included by integrating the result over its
  angular flux distribution.  The additional spatial-coherence assumptions
  needed to apply a plane-wave \(N^2\) form factor to a macroscopic sample
  are stated explicitly in Sec.~\ref{sec:coherence}.
\end{enumerate}

The external magnetic field defines the spin quantisation axis
\(\hat z\).  As in the main text, the spin-dependent interaction is
\begin{equation}
 H_{\rm int}
 =\Sigma_z J_z+\Sigma_-J_-+\Sigma_+J_+ ,
\end{equation}
where
\begin{equation}
\begin{aligned}
 \Sigma_z
 &=\sqrt{2}G_F g_A\sum_i
   \bar\nu_i\gamma^3(1-\gamma^5)\nu_i,\\
 \Sigma_\pm
 &=\frac{G_Fg_A}{\sqrt{2}}\sum_i
   \bar\nu_i(\gamma^1\mp{\rm i}\gamma^2)
   (1-\gamma^5)\nu_i .
\end{aligned}
\label{eq:Sigma}
\end{equation}
The upper sign denotes spin excitation and the lower sign spin
de-excitation.  The spin splitting is \(\omega_0>0\).

\section{Reactor flux}
\label{sec:flux}

Let
\begin{equation}
 \phi(E,\hat p)
 \equiv\frac{\dd^3\Phi_{\bar\nu_e}}
 {\dd E\,\dd\Omega_{\hat p}}
\label{eq:angularflux}
\end{equation}
be the differential antineutrino flux at the sample.  It is a physical
one-helicity flux; there is no average over two initial helicities.  For a
pointlike, isotropic \(^{235}{\rm U}\) source at distance \(L\),
\begin{equation}
 \phi(E,\hat p)
 =\frac{\dot N_f}{4\pi L^2}S_{235}(E)
   \delta^{(2)}(\hat p-\hat n_R),
\qquad
 \dot N_f=\frac{P_{\rm th}}{Q_f},
\label{eq:pointflux}
\end{equation}
where \(S_{235}(E)\) is the yield per unit energy per fission, \(P_{\rm th}\)
is the thermal power, and \(Q_f\) is the energy released per fission.  For
an extended core, Eq.~\eqref{eq:pointflux} is replaced by the sum or volume
integral of the contributions from each core element.  All subsequent
formulae remain valid with the general angular flux
Eq.~\eqref{eq:angularflux}.

A produced electron antineutrino is a superposition of mass eigenstates.
The neutral-current interaction preserves the mass index.  After summing
over the unobserved final mass eigenstate, the leading rate is proportional
to
\(\sum_i|U_{ei}|^2=1\).  Reactor oscillations therefore do not alter the
neutral-current rate in the limit \(m_i^2/E^2\to0\).  This statement would
not apply to flavour-dependent or charged-current couplings.

\section{Single-spin rate}
\label{sec:single}

\subsection{Kinematics}

Write the incoming and outgoing massless four-momenta as
\begin{equation}
 p^\mu=E(1,\hat p),
\qquad
 p'^\mu=E'(1,\hat p').
\end{equation}
With recoil neglected, energy conservation gives
\begin{equation}
\begin{aligned}
 E'_+&=E-\omega_0,
 & E&\geq\omega_0
 &&\text{(excitation)},\\
 E'_-&=E+\omega_0,
 & E&\geq0
 &&\text{(de-excitation)}.
\end{aligned}
\label{eq:Epm}
\end{equation}
The momentum transferred to the fixed target is
\(\vec q_\pm=\vec p-\vec p\,'_\pm\).

\subsection{Definite-helicity spinor traces}

For an incoming active reactor antineutrino, summing over the allowed
outgoing antineutrino state gives
\begin{equation}
\begin{aligned}
 {\cal S}_+
 &\equiv
 \sum_c
 \left|
 \bar v_+(\vec p)
 (\gamma^1-{\rm i}\gamma^2)(1-\gamma^5)
 v_c(\vec p\,')
 \right|^2,\\
 {\cal S}_-
 &\equiv
 \sum_c
 \left|
 \bar v_+(\vec p)
 (\gamma^1+{\rm i}\gamma^2)(1-\gamma^5)
 v_c(\vec p\,')
 \right|^2 .
\end{aligned}
\label{eq:tracedef}
\end{equation}
The subscript on \(v_+\) denotes the positive-helicity antineutrino, not a
spin excitation.  Using
\(\sum_c v_c(p')\bar v_c(p')=\slashed p'\) in the massless limit, the two
traces evaluate to
\begin{equation}
\boxed{
\begin{aligned}
 {\cal S}_+
 &=16EE'(1+\hat p\cdot\hat z)
          (1-\hat p'\cdot\hat z),\\
 {\cal S}_-
 &=16EE'(1-\hat p\cdot\hat z)
          (1+\hat p'\cdot\hat z).
\end{aligned}}
\label{eq:spinortraces}
\end{equation}
Equation~\eqref{eq:spinortraces} is the main difference from an isotropic,
helicity-independent bath.  The parity-violating weak interaction retains
the direction of the reactor beam relative to the quantisation axis.

\subsection{Golden-rule result}

Starting with the invariant phase-space measure
\(\dd\widetilde p=\dd^3p/[(2\pi)^3 2E]\), the antineutrino contribution is
\begin{equation}
\begin{aligned}
 \gamma_\pm={}&
 \pi G_F^2g_A^2
 \int\dd\widetilde p\,\dd\widetilde p'\,
 (2\pi)^3\frac{\dd^3n_{\bar\nu}}{\dd^3p}\,
 \delta(E'_\pm-E')\,{\cal S}_\pm .
\end{aligned}
\label{eq:fgrphase}
\end{equation}
For a relativistic beam,
\(\dd^3n_{\bar\nu}/(\dd E\,\dd\Omega)=\phi(E,\hat p)\).
Substitution of Eq.~\eqref{eq:spinortraces} then yields
\begin{equation}
\boxed{
\begin{aligned}
 \gamma_+
 &=\frac{G_F^2g_A^2}{2\pi^2}
   \int_{\omega_0}^{\infty}\dd E\int\dd\Omega_{\hat p}\,
   \phi(E,\hat p)(E-\omega_0)^2\\
 &\hspace{18mm}\times
   \int\dd\Omega_{\hat p'}
   (1+\hat p\cdot\hat z)(1-\hat p'\cdot\hat z),\\
 \gamma_-
 &=\frac{G_F^2g_A^2}{2\pi^2}
   \int_0^{\infty}\dd E\int\dd\Omega_{\hat p}\,
   \phi(E,\hat p)(E+\omega_0)^2\\
 &\hspace{18mm}\times
   \int\dd\Omega_{\hat p'}
   (1-\hat p\cdot\hat z)(1+\hat p'\cdot\hat z).
\end{aligned}}
\label{eq:singlegeneral}
\end{equation}
The normalisation in Eq.~\eqref{eq:singlegeneral} is for the physical
active-helicity flux.  In particular, it contains no factor \(1/2\) from
averaging over an unoccupied initial helicity.

For a pencil beam along \(\hat n_R\), define
\begin{equation}
 c\equiv\cos\alpha=\hat n_R\cdot\hat z,
\qquad
 \phi_R(E)\equiv\int\dd\Omega_{\hat p}\,\phi(E,\hat p).
\end{equation}
The unrestricted outgoing angular integral gives the compact result
\begin{equation}
\boxed{
\begin{aligned}
 \gamma_+
 &=\frac{2G_F^2g_A^2}{\pi}(1+c)
   \int_{\omega_0}^{\infty}\dd E\,
   \phi_R(E)(E-\omega_0)^2,\\
 \gamma_-
 &=\frac{2G_F^2g_A^2}{\pi}(1-c)
   \int_0^{\infty}\dd E\,
   \phi_R(E)(E+\omega_0)^2.
\end{aligned}}
\label{eq:singlepencil}
\end{equation}
Thus \(\gamma_+=0\) for a beam antiparallel to \(+\hat z\), while
\(\gamma_-=0\) for a beam parallel to \(+\hat z\), within the massless and
recoil-free approximations.  Reversing either the beam or magnetic-field
direction interchanges the two rates.

\subsection{\texorpdfstring{Small-\(\omega_0/E\) expansion}
{Small-omega0/E expansion}}

For a transverse beam, \(c=0\), define the flux moments
\begin{equation}
 I_n\equiv\int_0^\infty\dd E\,E^n\phi_R(E).
\end{equation}
The negligible interval \(0<E<\omega_0\) aside,
Eq.~\eqref{eq:singlepencil} becomes
\begin{equation}
\begin{aligned}
 \gamma_+&={\cal C}
 (I_2-2\omega_0 I_1+\omega_0^2I_0),\\
 \gamma_-&={\cal C}
 (I_2+2\omega_0 I_1+\omega_0^2I_0),
 \qquad
 {\cal C}\equiv\frac{2G_F^2g_A^2}{\pi}.
\end{aligned}
\label{eq:momentexpansion}
\end{equation}
If \(\gamma_0\equiv{\cal C}I_2\), the leading kinematic imbalance is
\begin{equation}
 \boxed{
 \frac{\gamma_--\gamma_+}{\gamma_0}
 =4\omega_0\frac{I_1}{I_2}
 +{\cal O}\left(\frac{\omega_0^2}{E^2}\right).}
\label{eq:imbalance}
\end{equation}
This replaces the nonrelativistic suppression
\(\sim2m_\nu\omega_0/p_\nu^2\).  For a non-transverse beam, the
\({\cal O}(1)\) factors \(1\pm c\) dominate over the minute kinematic
effect in Eq.~\eqref{eq:imbalance}.

\section{\texorpdfstring{\(^{235}\mathrm{U}\)}{235U} spectrum benchmark}
\label{sec:numeric}

The file \texttt{../235U\_nspec.csv} contains a piecewise-constant spectrum
\(S_{235}(E)=\dd N_{\bar\nu}/(\dd E\,\dd{\rm fission})\).  Defining
\begin{equation}
 M_n=\int\dd E\,E^nS_{235}(E),
\end{equation}
direct integration gives
\begin{equation}
\begin{aligned}
 M_0&=6.077743311~{\rm fission}^{-1},\\
 M_1&=8.815892881~{\rm MeV\,fission}^{-1},\\
 M_2&=21.54413023~{\rm MeV^2\,fission}^{-1}.
\end{aligned}
\label{eq:spectrummoments}
\end{equation}
The mean and root-mean-square energies are \(1.45052\) MeV and
\(1.88275\) MeV, respectively, while the energy controlling
Eq.~\eqref{eq:imbalance} is
\begin{equation}
 E_{\rm eff}\equiv\frac{M_2}{M_1}=2.44378~{\rm MeV}.
\end{equation}
For \(\omega_0=10^{-8}\) eV,
\begin{equation}
 4\omega_0\frac{M_1}{M_2}=1.6368\times10^{-14}.
\end{equation}

As an illustrative normalisation, take \(g_A=1/2\),
\(P_{\rm th}=1\) GW, \(L=10\) m, and \(Q_f=202.36\) MeV.  The point-source
flux and the spectrum-averaged transverse single-spin cross section are
\begin{equation}
\begin{aligned}
 \Phi_R&=1.49175\times10^{13}~{\rm cm}^{-2}{\rm s}^{-1},\\
 \vev{\sigma_\pm}
 &=2.98854\times10^{-44}~{\rm cm}^2 ,
\end{aligned}
\end{equation}
which give
\begin{equation}
 \gamma_+\simeq\gamma_-
 =4.45817\times10^{-31}~{\rm s}^{-1}
\label{eq:numericgamma}
\end{equation}
at the displayed precision.  These numbers scale as
\(g_A^2P_{\rm th}/L^2\).  They are not a claim about a particular reactor;
the power, distance, fission energy, coupling, and angle are exposed as
inputs in \texttt{reactor\_rates.py}.

\section{\texorpdfstring{\(N\)}{N}-spin ensemble}
\label{sec:ensemble}

For spins at positions \(\vec x_\alpha\), define
\begin{equation}
 F_\pm(\vec q)
 =\sum_{{\rm f}_{\rm spin}}
 \left|
 \left\langle{\rm f}_{\rm spin}\right|
 \sum_\alpha J_\pm^\alpha e^{-{\rm i}\vec q\cdot\vec x_\alpha}
 \left|{\rm i}_{\rm spin}\right\rangle
 \right|^2.
\label{eq:Fgeneral}
\end{equation}
Repeating the single-spin calculation without performing the outgoing
angular integral gives
\begin{equation}
\boxed{
\begin{aligned}
 \Gamma_+
 &=\frac{G_F^2g_A^2}{2\pi^2}
 \int_{\omega_0}^\infty\dd E\int\dd\Omega_{\hat p}\,
 \phi(E,\hat p)(E-\omega_0)^2\\
 &\quad\times\int\dd\Omega_{\hat p'}
 (1+\hat p\cdot\hat z)(1-\hat p'\cdot\hat z)
 F_+(\vec q_+),\\
 \Gamma_-
 &=\frac{G_F^2g_A^2}{2\pi^2}
 \int_0^\infty\dd E\int\dd\Omega_{\hat p}\,
 \phi(E,\hat p)(E+\omega_0)^2\\
 &\quad\times\int\dd\Omega_{\hat p'}
 (1-\hat p\cdot\hat z)(1+\hat p'\cdot\hat z)
 F_-(\vec q_-).
\end{aligned}}
\label{eq:ensemblegeneral}
\end{equation}
This is the reactor-source replacement for Eq.~(B.18) of the main
manuscript.  Unlike the isotropic expression, the weak angular kernel
cannot generally be pulled outside the form-factor integral.

For the initial transverse product state
\(\left|\rightarrow\right\rangle^{\otimes N}\), the continuum form factor
of a uniform sphere of radius \(R\) is
\begin{equation}
 F_\pm^P(q)
 =\frac{N}{2}
 +\frac{N^2}{4}
 \left[\frac{3j_1(qR)}{qR}\right]^2 .
\label{eq:sphereF}
\end{equation}
The \(N/2\) term is the incoherent contribution; the \(N^2\) term contains
the collective contribution.  For a pencil beam, let
\(\mu=\hat p\cdot\hat p'\).  Averaging over the outgoing azimuth about
\(\hat p\) gives
\(\vev{\hat p'\cdot\hat z}_{\rm az}=c\mu\), and hence
\begin{equation}
\boxed{
\begin{aligned}
 \Gamma_\pm
 =\frac{G_F^2g_A^2}{\pi}
 \int\dd E\,\phi_R(E)E_\pm'^2(1\pm c)
 \int_{-1}^{1}\dd\mu\,(1\mp c\mu)F_\pm(q_\pm),
\end{aligned}}
\label{eq:ensemblepencil}
\end{equation}
with the excitation threshold understood and
\begin{equation}
\begin{aligned}
 q_\pm^2
 &=E^2+E_\pm'^2-2EE_\pm'\mu\\
 &=\omega_0^2+2EE_\pm'(1-\mu).
\end{aligned}
\label{eq:qrel}
\end{equation}
For \(F_\pm=1\), Eq.~\eqref{eq:ensemblepencil} reduces to
Eq.~\eqref{eq:singlepencil}.

\section{Relativistic coherence limits}
\label{sec:coherence}

In the exactly forward direction,
\begin{equation}
 q_{\pm,\min}=\omega_0,
\label{eq:qmin}
\end{equation}
instead of the nonrelativistic result
\(q_{\min}\sim m_\nu\omega_0/p_\nu\).  At small scattering angle
\(\theta\),
\begin{equation}
 q_\pm^2\simeq\omega_0^2+EE_\pm'\theta^2.
\label{eq:qsmall}
\end{equation}
The forward collective region therefore requires
\begin{equation}
 \omega_0R\ll1,
\qquad
 \theta\lesssim\frac{1}{ER}.
\label{eq:cohconditions}
\end{equation}
For \(R=1\) cm, \(\omega_0=10^{-8}\) eV gives
\(\omega_0R=5.07\times10^{-4}\), so longitudinal coherence is possible.
Using the root-mean-square energy of the supplied spectrum,
\begin{equation}
 ER=9.54\times10^{10},
\qquad
 \theta_{\rm coh}\simeq1.05\times10^{-11}.
\end{equation}
Thus the whole solid angle is never coherent for a centimetre-scale
reactor target; only a very narrow forward cone contributes to the
\(N^2\) term.

For \(ER\gg1\), \(\omega_0R\ll1\), and a spherical sample, the angular
integral can be done asymptotically:
\begin{equation}
\begin{aligned}
 \int\dd\Omega_{\hat p'}
 \left[\frac{3j_1(qR)}{qR}\right]^2
 &\simeq\frac{18\pi}{E^2R^2}
 \int_0^\infty\frac{\dd x}{x}j_1^2(x)\\
 &=\frac{9\pi}{2E^2R^2},
\end{aligned}
\label{eq:Wangular}
\end{equation}
where
\(\int_0^\infty\dd x\,j_1^2(x)/x=1/4\).
Inside this cone \(\hat p'\simeq\hat p\), so both circular-current kernels
approach \(1-c^2\).  The leading collective cross sections are consequently
\begin{equation}
\boxed{
 \sigma_{{\rm coh},+}
 \simeq\sigma_{{\rm coh},-}
 \simeq
 \frac{9G_F^2g_A^2N^2}{16\pi R^2}(1-c^2).}
\label{eq:sigmacoherent}
\end{equation}
Here \(R\) is expressed in natural units.  The leading result is independent
of \(E\): the elementary \(E^2\) scaling is cancelled by the
\((ER)^{-2}\) forward-cone fraction.

The incoherent terms for the transverse product state are
\begin{equation}
 \sigma_{{\rm inc},\pm}
 \simeq
 \frac{NG_F^2g_A^2}{\pi}E^2(1\pm c).
\label{eq:sigmaincoherent}
\end{equation}
For \(c=0\), their spectrum-integrated ratio is
\begin{equation}
 \frac{\Gamma_{\rm coh}}{\Gamma_{\rm inc}}
 \simeq\frac{9N}{16R^2}\frac{M_0}{M_2}.
\label{eq:cohratio}
\end{equation}
As a plane-wave benchmark, a sphere with \(R=1\) cm and spin density
\(n_s=3\times10^{22}\ {\rm cm}^{-3}\) contains
\(N=1.25664\times10^{23}\) spins.  Equations~\eqref{eq:sigmacoherent}
and~\eqref{eq:cohratio} give
\begin{equation}
\begin{aligned}
 \sigma_{{\rm coh},\pm}
 &=1.4580\times10^{-20}\ {\rm cm}^2,\\
 \Gamma_{{\rm coh},\pm}
 &=2.1750\times10^{-7}\ {\rm s}^{-1},\\
 \frac{\Gamma_{\rm coh}}{\Gamma_{\rm inc}}
 &=7.7646
\end{aligned}
\end{equation}
for the illustrative \(1\)-GW, \(10\)-m, transverse-beam normalisation of
Sec.~\ref{sec:numeric}.  These are ensemble transition rates, not
single-spin rates.

Two qualifications are important.
\begin{enumerate}
  \item When \(c=\pm1\), the leading forward kernel \(1-c^2\) vanishes.
  One of the two rates vanishes identically in the massless limit, while
  the other receives only a subleading contribution from nonzero
  scattering angles.  Equation~\eqref{eq:sigmacoherent} must not be used
  to form a \(0/0\) ratio in this aligned geometry.
  \item The \(N^2\) benchmark assumes that the incident neutrino state is
  sufficiently coherent across the sample for the plane-wave phase in
  Eq.~\eqref{eq:Fgeneral} to apply.  A real reactor has a finite core,
  finite source distance, and neutrino wave packets.  If those effects
  limit the mutual coherence length below \(R\), Eq.~\eqref{eq:sphereF}
  must be replaced by the corresponding two-point source correlation
  function.  The single-spin result and the incoherent \(N/2\) term are
  unaffected.  The numerical \(N^2\) result above should therefore be
  read as the plane-wave upper benchmark, not yet as a realistic reactor
  prediction.
\end{enumerate}

\section{Reproducibility}

The numerical sources are stored alongside this document:
\begin{itemize}
  \item \texttt{check\_spinor\_traces.py} constructs explicit Dirac gamma
  matrices and verifies Eq.~\eqref{eq:spinortraces} for \(10^4\) random
  direction pairs.
  \item \texttt{reactor\_rates.py} integrates
  \texttt{../235U\_nspec.csv}, evaluates the flux moments, single-spin
  rates, coherence angle, and the plane-wave ensemble benchmark.  Its
  power, distance, \(Q_f\), \(g_A\), \(\omega_0\), beam angle, radius, and
  spin density are command-line options.
\end{itemize}

\end{document}
